#   Thesis script, alpha diversity       #
#      Anna Maria Mavroselinou           #
##########################################

```{r cars}

EEZ_euphotic_heter <-  ps_protist_heter %>%  subset_samples(photic_zone== "Euphotic")

EEZ_euphotic_autot <-  ps_protist_autot %>%  subset_samples(photic_zone== "Euphotic")

STW_euphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Euphotic") 

STF_euphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Euphotic") 

SAW_euphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Euphotic") 

```

# Creating lists for some specific cases - WATER MASSES - APHOTIC


```{r pressure, echo=FALSE}

EEZ_aphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Aphotic")

STW_aphotic <- ps_protist_heter %>%  subset_samples(photic_zone== "Aphotic") 

STF_aphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Aphotic") 

SAW_aphotic <-  ps_protist_heter %>%  subset_samples(photic_zone== "Aphotic") 


```

```{r}
#### WATER MASS  & AREA DIVERSITY

#FIGURE 3
## EUPHOTIC - Water mass - FIGURE 

# Remove samples with NA in 'Area'
ps_protist_heter_clean <- subset_samples(ps_protist_heter, !is.na(Area))
ps_protist_heter_clean <- subset_samples(ps_protist_heter_clean, !is.na(Season))
ps_protist_heter_clean <- subset_samples(ps_protist_heter_clean, !is.na(photic_zone))

# Use euphotic or all (euphotic + aphotic)

CairoPDF(file = "Protist.Diversity_HET_Water.Mass", width = 10, height = 5) 
    plot_richness(ps_protist_heter_clean, x="Water_mass", color="Water_mass", measures = c("Observed","Shannon"), title = "Protist Heterotrophics") + geom_boxplot(alpha= 0.8, outlier.size = 0.2, position=position_dodge(.9)) +
        geom_point(size = 0.1) +
          #scale_color_manual(values = Water_mass_colors) +
        theme_bw() +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
dev.off()

```


```{r}
## BPM Euphotic - SEASONAL ##

#Remove samples with NA in 'Season'
ps_protist_heter <- subset_samples(ps_protist_heter, !is.na(Season))
#EEZ_euphotic_heter <- subset_samples(EEZ_euphotic_heter, !is.na(Area))
ps_protist_heter <- subset_samples(ps_protist_heter, !is.na(photic_zone))

#CairoPDF(file = "R.Figures/Richness/Protist.Diversity_Euphotic_SEASONAL", width = 10, height = 5) 
   v <-  plot_richness( ps_protist_heter_clean, x="Water_mass", color="Season", measures = c("Observed","Shannon"), title = "Protist HET - SEASONAL") + geom_boxplot(alpha= 0.8, outlier.size = 0.2, position=position_dodge(.9)) +
        geom_point(size = 0.1) +
         # scale_color_manual(values = Season_colors) +
        theme_bw() +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
#dev.off()

v
v$layers <- v$layers[-1]
v
v$layers <- v$layers[-2]
v

ggsave("Protist.Diversity_het_SEASONAL.pdf", dpi = 300,  width = 10, height = 5)

```

```{r}
# Use euphotic or all (euphotic + aphotic)


## BPM Euphotic - SEASONAL ##

#CairoPDF(file = "R.Figures/Richness/Protist.Diversity_Euphotic_SEASONAL", width = 10, height = 5) 
v <-  plot_richness( ps_protist_heter_clean, x="Water_mass", color="photic_zone", measures = c("Observed","Shannon"), title = "Protist HET -  Photic zone") + geom_boxplot(alpha= 0.8, outlier.size = 0.2, position=position_dodge(.9)) +
        geom_point(size = 0.1) +
         # scale_color_manual(values = Season_colors) +
        theme_bw() +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
#dev.off()

v
v$layers <- v$layers[-1]
v
v$layers <- v$layers[-2]
v

ggsave("Protist.Diversity_het_photic_zone.pdf", dpi = 300,  width = 10, height = 5)

```

```{r}
# Get the sample metadata
meta <- sample_data(ps_protist_heter) %>% data.frame()

# Check the unique values and counts in the depth zone column
table(meta$photic_zone)
#Aphotic Euphotic 
#  119      547 

v <- plot_richness( ps_protist_heter, x="photic_zone", color="photic_zone", measures = c("Observed","Shannon"), title = "Protist Heterotrophics") + geom_boxplot(alpha= 0.8, outlier.size = 0.2, position=position_dodge(.9)) +
        geom_point(size = 0.1) +
          #scale_color_manual(values = Water_mass_colors) +
        theme_bw() +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))

v
v$layers <- v$layers[-1]
v
v$layers <- v$layers[-2]
v

ggsave("Protist.Diversity_HET_photic.pdf", dpi = 300,  width = 10, height = 5)
 
```
```{r}
# Pull tax_table as dataframe
tax_df <- as.data.frame(tax_table(ps_protist))

# Get OTUs for each trophic mode
auto_otus <- rownames(tax_df[tax_df$trophic_mode == "photosynthetic", ])
hetero_otus <- rownames(tax_df[tax_df$trophic_mode == "heterotrophic", ])
mixo_otus <- rownames(tax_df[tax_df$trophic_mode == "mixotrophic", ])

# Subset the full phyloseq object
ps_autotroph <- prune_taxa(auto_otus, ps_protist)
ps_heterotroph <- prune_taxa(hetero_otus, ps_protist)
ps_mixotroph <- prune_taxa(mixo_otus, ps_protist)

# Alpha diversity for autotrophs
alpha_auto <- estimate_richness(ps_autotroph, measures = c("Shannon")) %>%
  rownames_to_column("file_code") %>%
  mutate(trophic_mode = "photosynthetic")

# Alpha diversity for heterotrophs
alpha_hetero <- estimate_richness(ps_heterotroph, measures = c("Shannon")) %>%
  rownames_to_column("file_code") %>%
  mutate(trophic_mode = "heterotrophic")

# Alpha diversity for mixotrophs
alpha_mixo <- estimate_richness(ps_mixotroph, measures = c("Shannon")) %>%
  rownames_to_column("file_code") %>%
  mutate(trophic_mode = "mixotrophic")

# Combine
alpha_combined <- bind_rows(alpha_auto, alpha_hetero,alpha_mixo)


meta_df <- data.frame(sample_data(ps_protist)) %>%
  rownames_to_column("file_code") %>%
  select(file_code, Water_mass, photic_zone)

# Join metadata
alpha_final <- left_join(alpha_combined, meta_df, by = "file_code")


ggplot(alpha_final %>% filter(photic_zone == "Euphotic"),
       aes(x = Water_mass, y = Shannon, color = trophic_mode)) +
  geom_boxplot(alpha = 0.75) +
  stat_summary(fun = mean, geom = "point", shape = 21, color = "black", size = 2,
               position = position_dodge(width = 0.75)) +
  labs(title = "Shannon Diversity (Euphotic Only)",
       x = "Water Mass", y = "Shannon Index", fill = "Trophic Mode") +geom_boxplot(alpha= 0.8, outlier.size = 0.5)+
  scale_fill_manual(values = c("photosynthetic" = "#1b9e77", "heterotrophic" = "#d95f02","mixotrophic" = "#7570b3")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Save plot
ggsave("Shannon_ALL_TrophicMode_WaterMass_eup.pdf", width = 12, height = 6, dpi = 300)



ggplot(alpha_final,
       aes(x = Water_mass, y = Shannon, color = trophic_mode)) +
  geom_boxplot(alpha = 0.75) +
  stat_summary(fun = mean, geom = "point", shape = 21, color = "black", size = 2,
               position = position_dodge(width = 0.75)) +
  labs(title = "Shannon Diversity (ALL DEPTHS)",
       x = "Water Mass", y = "Shannon Index", fill = "Trophic Mode") +geom_boxplot(alpha= 0.8, outlier.size = 0.5)+
  scale_fill_manual(values = c("photosynthetic" = "#1b9e77", "heterotrophic" = "#d95f02","mixotrophic" = "#7570b3")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#FIGURE 5
#Covariation
```{r}
# Reshape to wide format: one Shannon column per trophic mode
alpha_final_wide <- alpha_final %>%
  select(file_code, Shannon, trophic_mode) %>%
  pivot_wider(names_from = trophic_mode, values_from = Shannon,
              names_prefix = "Shannon_")


#Heterotrophic vsPhotosynthetic
cor.test(alpha_final_wide$Shannon_heterotrophic, alpha_final_wide$Shannon_photosynthetic, method = "pearson")

#Pearson's product-moment correlation

#data:  alpha_final_wide$Shannon_heterotrophic and alpha_final_wide$Shannon_photosynthetic
#t = 19.877, df = 669, p-value < 2.2e-16
#alternative hypothesis: true correlation is not equal to 0
#95 percent confidence interval:
# 0.5594581 0.6548314
#sample estimates:
#      cor 
#0.6093441 

#Heterotrophic vs Mixotrophic
cor.test(alpha_final_wide$Shannon_heterotrophic, alpha_final_wide$Shannon_mixotrophic, method = "pearson")

#	Pearson's product-moment correlation
#data:  alpha_final_wide$Shannon_heterotrophic and alpha_final_wide$Shannon_mixotrophic
#t = 26.207, df = 669, p-value < 2.2e-16
#alternative hypothesis: true correlation is not equal to 0
#95 percent confidence interval:
# 0.6722677 0.7471781
#sample estimates:
#      cor 
#0.7117406 

#Photosynthetic vs mixotrophic
cor.test(alpha_final_wide$Shannon_photosynthetic, alpha_final_wide$Shannon_mixotrophic, method = "pearson")

#	Pearson's product-moment correlation
#data:  alpha_final_wide$Shannon_photosynthetic and alpha_final_wide$Shannon_mixotrophic
#t = 26.228, df = 669, p-value < 2.2e-16
#alternative hypothesis: true correlation is not equal to 0
#95 percent confidence interval:
# 0.6725743 0.7474252
#sample estimates:
#      cor 
#0.7120166

# Subset only Shannon columns
shannon_df <- alpha_final_wide[, c("Shannon_heterotrophic", "Shannon_mixotrophic", "Shannon_photosynthetic")]

# Compute correlation matrix
cor_matrix <- cor(shannon_df, method = "pearson", use = "complete.obs")

# Rename row and column names for clarity in the plot
colnames(cor_matrix) <- c("Heterotrophic", "Mixotrophic", "Photosynthetic")
rownames(cor_matrix) <- c("Heterotrophic", "Mixotrophic", "Photosynthetic")

# Plot
# Save the correlation plot as a PNG file
png("Pairwise.png", width = 12, height = 7, units = "in", res = 300, bg = "white")
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", addCoef.col = "black", 
         number.cex = 0.8, 
         col = colorRampPalette(c("blue", "white", "red"))(200))
dev.off()


# Only required columns
shannon_df <- alpha_final_wide[, c("Shannon_heterotrophic", "Shannon_mixotrophic", "Shannon_photosynthetic")]


# Plot pairwise scatterplots
ggpairs(shannon_df, 
        lower = list(continuous = wrap("smooth", method = "lm", se = FALSE, color = "steelblue")),
        upper = list(continuous = wrap("cor", size = 4)),
        diag = list(continuous = wrap("densityDiag", alpha = 0.5)))

```



#############################  continuous data to discrete ###################################
“How does heterotrophic protist diversity in the euphotic zone change as we move from south to north (latitude), and is this change related to chlorophyll levels?”

```{r}

#I will extracting the sample data from the phyloseq object, create the Ranking columns and then import the new coluns in the phyloseq object 
Euphotic_ranks <- sample_data(EEZ_euphotic_heter) %>%
  data.frame() %>%
  mutate(
    RankLat1 = cut(Lat, breaks = c(-80, -75, -70, -65, -60, -55, -50, -45, -40, -35, -30),
                   labels = c("-80","-70", "-65", "-60", "-55", "-50", "-45", "-40", "-35", "-30"),
                   include.lowest = TRUE),
    RankLat2 = ntile(Lat, 20),
    RankChla = ntile(Chla_total, 4)) %>%
  select(Samples, RankLat1, RankLat2,RankChla) %>%
  distinct(Samples, .keep_all = TRUE)

#Adding the new columns in the metadata
meta_orig <- sample_data(EEZ_euphotic_heter) %>% data.frame()

Euphotic_ranks$RankChla=as.factor(Euphotic_ranks$RankChla)
Euphotic_ranks$RankLat1=as.factor(Euphotic_ranks$RankLat1)
Euphotic_ranks$RankLat2=as.factor(Euphotic_ranks$RankLat2)

meta_merged <- left_join(meta_orig, Euphotic_ranks, by = "Samples")
rownames(meta_merged) <- meta_merged$Samples 

sample_data(EEZ_euphotic_heter) <- sample_data(meta_merged)



# Check value ranges for each chlorophyll rank
sample_data(EEZ_euphotic_heter) %>%
  data.frame() %>%
  group_by(RankChla) %>%
  summarise(
    Min_Chla = min(Chla_total, na.rm = TRUE),
    Max_Chla = max(Chla_total, na.rm = TRUE),
    Mean_Chla = mean(Chla_total, na.rm = TRUE),
    n_samples = n()
  )



# Extract sample data as dataframe
meta_df <- sample_data(EEZ_euphotic_heter) %>%
  data.frame()

# Add sample names as a column
meta_df$Samples <- rownames(meta_df)

# Reorder if you want Samples first (optional)
meta_df <- meta_df %>%
  select(Samples, everything())

# Write to Excel
write.xlsx(meta_df, file = "EEZ_euphotic_heter_metadata.xlsx", rowNames = FALSE)
# convert integer to factor and levels
#Euphotic_rankss$RankChla=as.factor(Euphotic_rankss$RankChla)
#Euphotic_rankss$RankNO3=as.factor(Euphotic_rankss$RankNO3) 
#Euphotic_rankss$RankDRSi=as.factor(Euphotic_rankss$RankDRSi) 


```

How protist heterotroph diversity in the euphotic zone changes along latitude and potentially relates to chlorophyll concentration?
```{r}
#CairoPDF(file = "R.Figures/Richness/Protist.Diversity_EUPHOTIC_Latitude1_Area2_March2022", width = 10, height = 5) 

EEZ_euphotic_heter <- subset_samples(EEZ_euphotic_heter, !is.na(RankLat1))
EEZ_euphotic_heter <- subset_samples(EEZ_euphotic_heter, !is.na(RankChla))

# Extract original metadata
meta_orig <- sample_data(ps_protist_heter) %>% data.frame()

# Merge by 'Samples'
meta_merged <- left_join(meta_orig, Euphotic_ranks, by = "Samples")

# Set row names back for phyloseq compatibility
rownames(meta_merged) <- meta_merged$Samples

# Add back to phyloseq object
sample_data(ps_protist_heter) <- sample_data(meta_merged)

# If 'DepthLayer' or 'Zone' column marks euphotic samples
ps_euphotic_heter <- subset_samples(ps_protist_heter, photic_zone == "Euphotic")


# Estimate richness measures
richness_df <- estimate_richness(ps_euphotic_heter, measures = c("Observed", "Shannon"))

# Add Sample names as a column
richness_df$Samples <- rownames(richness_df)

# dont know why the '-' in the rownames are changed into '.' 
richness_df$Samples <- chartr(".", "-", richness_df$Samples) # changed the naming back so it can be joined with sample_ID from medadata
rownames(richness_df) <- richness_df$Samples 

# Extract sample data
meta_df <- sample_data(ps_euphotic_heter) %>% data.frame()

# Merge
richness_df <- richness_df %>%
  left_join(meta_df, by = "Samples")

# Find the row with the highest Shannon diversity
richness_df[which.max(richness_df$Shannon), ]


  t <- plot_richness(EEZ_euphotic_heter, x="RankLat1", color="RankChla", measures = c("Observed","Shannon"), title = "Protist Heter - Euphotic - Latitude") + geom_boxplot(alpha= 0.8, outlier.size = 0.2) +
               # scale_color_manual(values = Area_colors) +
        geom_point(size = 0.1) +
        theme_bw() +
        #scale_x_continuous(limits = c(0, 10)) +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
t

# Save plot
ggsave("Euph_lat_chlor.pdf", width = 12, height = 6, dpi = 300)

#FIGURE 2
#With lat on x and water mass at y
  t <- plot_richness(EEZ_euphotic_heter, x="RankLat1", color="Water_mass", measures = c("Observed","Shannon"), title = "Protist Heter - Euphotic - Latitude") + geom_boxplot(alpha= 0.8, outlier.size = 0.2) +
               # scale_color_manual(values = Area_colors) +
        geom_point(size = 0.1) +
        theme_bw() +
        #scale_x_continuous(limits = c(0, 10)) +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
t

# Save plot
ggsave("Euph_lat_water.pdf", width = 12, height = 6, dpi = 300)



#--------------------------------------------------------------------------------------------
#FIGURE 4
#Using rank latitudes or discrete water masses in the X axis and rank chla concentrations within each of this.         
richness_df$Lat_Chla_Group <- paste(richness_df$RankLat1, richness_df$RankChla, sep = "_")

sample_data(EEZ_euphotic_heter)$Lat_Chla_Group <- with(sample_data(EEZ_euphotic_heter), paste(RankLat1, RankChla, sep = "_"))

#Compute ranges per rank
meta_df <- sample_data(EEZ_euphotic_heter) %>% data.frame()

# If Chla_total is a character, convert safely to numeric:
meta_df$Chla_total <- as.numeric(as.character(meta_df$Chla_total))

# Compute Chl a ranges
rank_labels <- meta_df %>%
  group_by(RankChla) %>%
  summarise(
    MinChla = round(min(Chla_total, na.rm = TRUE), 2),
    MaxChla = round(max(Chla_total, na.rm = TRUE), 2)
  ) %>%
  mutate(
    RankChla_label = paste0("Rank ", RankChla, ": ", MinChla, "–", MaxChla, " µg/L")
  )
 
#Map the new labels to your metadata
# Create a named vector to relabel
rank_label_map <- setNames(rank_labels$RankChla_label, rank_labels$RankChla)

# Relabel directly in metadata
meta_df$RankChlaLabel <- rank_label_map[as.character(meta_df$RankChla)]

# Add back to phyloseq
sample_data(EEZ_euphotic_heter) <- sample_data(meta_df)

t <- plot_richness(EEZ_euphotic_heter, x="Lat_Chla_Group", color="RankChlaLabel", measures = c("Observed","Shannon"), title = "Protist Heter - Euphotic - Latitude") + geom_boxplot(alpha= 0.8, outlier.size = 0.2) +
               # scale_color_manual(values = Area_colors) +
        geom_point(size = 0.1) +
        theme_bw() +
        #scale_x_continuous(limits = c(0, 10)) +
        theme(axis.text.y = element_text(size = 12)) +
        theme(axis.title.y = element_text(size=12)) +
        theme(axis.title.x = element_text(size=12)) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 12))   +
        theme(legend.title = element_text(color = "blue", size = 12), legend.text = element_text(color = "black", size = 10))
t + labs(color = "Chlorophyll a Concentration \nRanks (µg/L ranges)") + labs(x = "Latitude Groups (°S) by Chlorophyll a Rank")

ggsave("Euph_lat_wchla_nested.pdf", width = 12, height = 6, dpi = 300)
